Avoiding tipping points in the management of ecological systems: a non-parametric Bayesian approach

### 2014-05-06 manuscripts/manuscript.Rmd
update helping scripts

toggle = "markup" # results argument
```

@@

@@

We use a Bayesian nonparameteric regression approach based on Gaussian process priors. 
The use of Gaussian process (GP) regression (or "kriging" in the geospatial
literature) to formulate a predictive model is relatively new in the
context of modeling dynamical systems [@Kocijan2005], and was first introduced
in the context ecological modeling and fisheries management in @Munch2005.
GP models have subsequently been used to test for the presence of Allee
effects [@SugenoMunch], estimate the maximum reproductive
rate [@SugenoMunch], determine temporal variation in food
availability [@Sigourney], and provide a basis for identifying
model-mispecification [@Thorson].
An accessible and thorough introduction to the formulation and use of
GPs can be found in @Rasmussen2006. The nonparametric monicker for
the GP approach is unfortunate as (a) it still involves the estimation
of parameters, and (b), statisticians also use non-parametric to mean
models that make only weak distributional assumptions.  (@signTestREF). 
However, we retain the terminology for consistency with previous
studies.

@@

We infered posterior distributions for the parameters of each model
in a Bayesian context using Gibbs sampling (implemented in R [@RTeam]
using jags, [@R2jags]).  We choose uninformative uniform priors for all
parameters (See Appendix, Figures and tables, and the R code
provided). One-step-ahead predictions of these model fits are shown in
Figure 1.  Each chain was tested for Gelman-Rubin convergence and results wer robust
to longer runs.  For each simulation we also applied several commonly
used model selection criteria (AIC, BIC, DIC, see @BurnhamAndersonREF)
to identify the best fitting model.

@@

These results are not sensitive to the modeling details of the simulation.  The GPDP estimate remains very close to the optimal solution obtained by knowning the true model across changes to the training simulation, noise scale, parameters or structure of the underlying model, as seen in Figure 7. Results are pooled across different random seeds (111, 222, 333), noise values of 0.01 and 0.05 (not large enough to violate the self-sustaining criterion of @Reed1979), and different randomly generated parameters sets for each model, and using either the Myers or Allen as the underlying structure.  While the GPDP performs slightly closer to optimal on the smaller noise levels, and closer on the Allen model then the Myers model, the differences are small compared to the values observed in using the wrong parametric model.  The GPDP solution consistently avoids either failing to harvest at all or crashing the stock, which results in the very low profits/values seen in parametric models in the previous figure. The sensitivity analysis is compared against the optimal solution alone and not shown against the estimates of the various parameteric models (but see Figure 6), where convergence of the Gibbs sampling could contribute to their poorer performance if not checked by hand for each configuration.

![Sensitivity Analysis.  Histogram shows the ratio of the realized net present value / profits derived when managing under the GPDP Gaussian Process approach over the optimal value given the true model and true parameters. Values near 1 indicate nearly optimal performance. Simulations cover a range of underlying parameter, model, replicate, and noise levels.](figure/sensitivity.pdf)

@@

Lastly, it should be noted that outside the data, the GP reverts to
the prior, and consequently the choice of the prior can also play a
significant role in determining the optimal policy inferred by the
SDP.  In the examples shown here we have selected a prior that is both
relatively uninformative (due to the broad priors placed on its parameters
$\ell$ and $\sigma$ and simple (mean zero, radial basis function kernel).
In practice, both the choice of mean and the choice of the covariance
function may be chosen to confer particular biological properties,
as well as more biologically informed priors for $\ell$ and $\sigma$.
In principle, this may allow a manager to improve the performance of the
GPDP approach by adding only enough additional detail as is justified.
For instance, it would be possible to use a linear or a Ricker-shaped
mean in the prior without making the much stronger assumption that the
Ricker is the structurally correct model [@SugenoMunch]. Future research
should focus on identifying criteria to ensure the prior and the value function
are chosen appropriately for the problem at hand.

>>>>

toggle = "hide" # results argument
```

@@

```{r sensitivity-calc}
source("components/sensitivity.R")
```

@@

We use a Bayesian nonparameteric regression approach based on Gaussian process priors. 
The use of Gaussian process (GP) regression (or "kriging" in the geospatial
literature) to formulate a predictive model is relatively new in the
context of modeling dynamical systems [@Kocijan2005], and was first introduced
in the context ecological modeling and fisheries management in @Munch2005.
GP models have subsequently been used to test for the presence of Allee
effects [@Sugeno2013], estimate the maximum reproductive
rate [@Sugeno2013], determine temporal variation in food
availability [@Sigourney], and provide a basis for identifying
model-mispecification [@Thorson2014].
An accessible and thorough introduction to the formulation and use of
GPs can be found in @Rasmussen2006. The nonparametric monicker for
the GP approach is unfortunate as (a) it still involves the estimation
of parameters, and (b), statisticians also use non-parametric to mean
models that make only weak distributional assumptions  [e.g. @Lehmann1975]. 
However, we retain the terminology for consistency with previous
studies.

@@

We infered posterior distributions for the parameters of each model
in a Bayesian context using Gibbs sampling (implemented in R [@RTeam]
using jags, [@R2jags]).  We choose uninformative uniform priors for all
parameters (See Appendix, Figures and tables, and the R code
provided). One-step-ahead predictions of these model fits are shown in
Figure 1.  Each chain was tested for Gelman-Rubin convergence and results wer robust
to longer runs.  For each simulation we also applied several commonly
used model selection criteria (AIC, BIC, DIC, see @Burnham2002)
to identify the best fitting model.

@@

These results are not sensitive to the modeling details of the simulation.
The GPDP estimate remains very close to the optimal solution obtained
by knowning the true model across changes to the training simulation,
noise scale, parameters or structure of the underlying model, as seen in
Figure 7. The sensitivity analysis uses a factorial design of eight replicates 
of three different randomly generated parameter sets for each of two different 
generating models (Allen and Myers) over two different noise levels 
(0.01 and 0.05), for a total of 8 x 3 x 2 x 2 = 96 scenarios. In Figure 7, 
results are pooled across different random seeds (111, 222,
333), noise values of 0.01 and 0.05 (not large enough to violate the
self-sustaining criterion of @Reed1979), and different randomly generated
parameters sets for each model, and using either the Myers or Allen as the
underlying structure.  While the GPDP performs slightly closer to optimal
on the smaller noise levels, and closer on the Allen model then the Myers
model, the differences are small compared to the values observed in using
the wrong parametric model.  The GPDP solution consistently avoids either
failing to harvest at all or crashing the stock, which results in the very
low profits/values seen in parametric models in the previous figure. The
sensitivity analysis is compared against the optimal solution alone and
not shown against the estimates of the various parameteric models (but
see Figure 6), where convergence of the Gibbs sampling could contribute
to their poorer performance if not checked by hand for each configuration.

```{r sensitivity, dependson="sensitivity-calc", echo=FALSE, fig.cap="Sensitivity Analysis.  Histogram shows the ratio of the realized net present value / profits derived when managing under the GPDP Gaussian Process approach over the optimal value given the true model and true parameters. Values near 1 indicate nearly optimal performance. Simulations cover a range of underlying parameter, model, replicate, and noise levels."}
ggplot(dat) + geom_histogram(aes(value, fill=noise)) + xlim(0,1.0) + theme_bw() + xlab("value as fraction of the optimal") + facet_wrap(~model, ncol=1)
```

@@

Lastly, it should be noted that outside the data, the GP reverts to
the prior, and consequently the choice of the prior can also play a
significant role in determining the optimal policy inferred by the
SDP.  In the examples shown here we have selected a prior that is both
relatively uninformative (due to the broad priors placed on its parameters
$\ell$ and $\sigma$ and simple (mean zero, radial basis function kernel).
In practice, both the choice of mean and the choice of the covariance
function may be chosen to confer particular biological properties,
as well as more biologically informed priors for $\ell$ and $\sigma$.
In principle, this may allow a manager to improve the performance of the
GPDP approach by adding only enough additional detail as is justified.
For instance, it would be possible to use a linear or a Ricker-shaped
mean in the prior without making the much stronger assumption that the
Ricker is the structurally correct model [@Sugeno2013]. Future research
should focus on identifying criteria to ensure the prior and the value function
are chosen appropriately for the problem at hand.
